shrimp_goby_boatnoise

### 2025-08-21 ms_boatnoise_shrimp_goby.Rmd
Updated most documents in preperation for publication. Formatting changes and additional analyses added.

**SUMMARY**  
Human activities are reshaping marine soundscapes globally, with persistent low-frequency noise from boat traffic recognised as a primary driver of marine noise pollution. Despite growing concern, empirical data for most marine life, in natural settings, remains sparse, particularly for fish and crustaceans whose auditory ranges overlap with the frequencies of boat noise. Accordingly, we conducted a field experiment to evaluate how boat noise from different engine types (2-stroke and 4-stroke) affects the mutualistic partnership between Steinitz’s goby (Amblyeleotris steinitzi) and snapping shrimp (Alpheus spp.). Across 123 mutualistic partnerships, we recorded behavioural responses before, during, and after noise exposure. Both species increased burrow use during noise exposure, indicating heightened risk perception, yet their tactile communication—critical for the functioning of the mutualism—remained unaltered. Notably, gobies responded more strongly to 4-stroke boat noise, reducing time outside their burrows by 22%, while shrimp were more sensitive to 2-stroke noise, reducing burrow use by 34%, reflecting species-specific sensitivities to different noise spectra. Behavioural changes were transient, rapidly reverting to baseline levels after the noise exposure stopped. These findings highlight divergent vulnerabilities tied to different engine acoustics and emphasise the need for targeted research to inform strategies for mitigating marine noise pollution.

@@

**CONTRIBUTOR ROLES**  
*🚧 to be added 🚧*

@@

**Licenses/restrictions placed on the data:** CC-BY 4.0

**Link to the associated publication:**     
🚧 *To be added* 🚧

**Recommended citation for this data:**      
🚧 *To be added based on OSF DOI* 🚧

@@

🌐 **Web:** *🚧 to be added 🚧*

@@

```{r}
#transforming the data frame into wide format and turning time spent in each zone into a proportion of the total time (i.e. a value between 0 and 1)
wide_data <- workingdf5 |>
  pivot_wider(
    id_cols = c(burrow_species, burrow_species_stage, burrow, stage, species, boat_id, stroke, site, number_conspecifics, shrimp_species),
    names_from = behaviour,
    values_from = mean_duration
  )

@@

final_df <- inner_join(goby_df, shrimp_df, by = "burrow_stage") |> 
  mutate(shrimp_contact = ifelse(is.na(shrimp_contact), ifelse(goby_out == 1, 0, 1 - goby_out), shrimp_contact)) |> 
  filter(site != "SITE O")

write_csv(final_df, "./1-data/final_df.csv")
```

@@

```{r}
get_prior(
  zoib_formula,
  data = final_df
) %>% 
  gt()
```

@@

#Now we run and save the model. **unhash** the code below if you haven't downloaded the saved model 'goby_out_model.rds' in the "3-models" folder.

# prior_model <- brm(
#   prior_formula,
#   data = final_df,
#   family = zero_one_inflated_beta(),
#   prior = priors,
#   control = list(adapt_delta = 0.99,
#                  max_treedepth = 12),
#   chains = 4, iter = 1000, warmup = 100,
#   cores = 4, threads = threading(2),
#   backend = "cmdstanr",
#   seed = 1,
#   sample_prior = "only",
#   file = "./3-models/prior_check"
# )

prior_model <-  readRDS(file = "./3-models/goby_out_model.rds")

@@

⏭️ **Skip this step** if you have downloaded the saved model 'goby_out_model.rds'. You will load the model in the next step.

```{r}
# goby_out_model <- brm(
#   goby_out_formula_1,
#   data = final_df,
#   family = zero_one_inflated_beta(),
#   prior = priors,
#   control = list(adapt_delta = 0.99,
#                  max_treedepth = 12),
#   chains = 4, iter = 4000, warmup = 1000,
#   cores = 4, threads = threading(2),
#   backend = "cmdstanr",
#   seed = 1,
#   file = "./3-models/goby_out_model"
# )
```

Here we reload the saved model.

```{r}
goby_out_model <-  readRDS(file = "./3-models/goby_out_model.rds")
```

@@

```{r}
goby_out_plot3
ggsave("./2-figures/goby_out_plot3.jpeg", plot = goby_out_plot3, units = "px", width = 2200, height = 1400, dpi = 300)
ggsave("./2-figures/goby_out_plot3.pdf", plot = goby_out_plot3, units = "px", width = 2200, height = 1400)
```

@@

⏭️ **Skip this step** if you have downloaded the saved model 'shrimp_out_model.rds'. You will load the model in the next step

```{r}
# shrimp_out_model <- brm(
#   shrimp_out_formula,
#   data = final_df,
#   family = zero_one_inflated_beta(),
#   prior = priors,
#   control = list(adapt_delta = 0.99,
#                  max_treedepth = 12),
#   chains = 4, iter = 4000, warmup = 1000,
#   cores = 4, threads = threading(2),
#   backend = "cmdstanr",
#   seed = 1,
#   file = "./3-models/shrimp_out_model"
# )
```

Here we reload the model.

```{r}
shrimp_out_model <-  readRDS(file = "./3-models/shrimp_out_model.rds")
```

@@

shrimp_out_plot3
ggsave("./2-figures/shrimp_out_plot3.jpeg", plot = shrimp_out_plot3, units = "px", width = 2200, height = 1400, dpi = 300)
ggsave("./2-figures/shrimp_out_plot3.pdf", plot = shrimp_out_plot3, units = "px", width = 2200, height = 1400)
```

@@

⏭️ **Skip this step** if you have downloaded the saved model 'shrimp_contact_model.rds'. You will load the model in the next step.

```{r}
# shrimp_contact_model <- brm(
#   bf(shrimp_contact ~ stroke*stage + shrimp_number + goby_number + shrimp_species + (1|site/burrow),
#      phi ~ stroke*stage + shrimp_number + goby_number + shrimp_species + (1|site/burrow),
#      zoi ~ stroke*stage + shrimp_number + goby_number + shrimp_species + (1|site/burrow),
#      coi ~ stroke*stage + shrimp_number + goby_number + shrimp_species + (1|site/burrow)
#      ),
#   data = final_df,
#   prior = priors,
#   family = zero_one_inflated_beta(),
#   control = list(adapt_delta = 0.99, max_treedepth = 12),
#   chains = 4, iter = 6000, warmup = 1000, seed = 1, backend = "cmdstanr", cores = 4, threads = threading(2),
#   file = "./3-models/shrimp_contact_model"
# )
```

Here we reload the model.

```{r}
shrimp_contact_model <-  readRDS(file = "./3-models/shrimp_contact_model.rds")
```

@@

shrimp_contact_model_estimates_two <- shrimp_contact_model_estimates |> 
  dplyr::filter(stroke == "2")|> 
  rename(.value = emmean)
```
#### Plotting model estimates and raw data

shrimp_contact_model_draws <- shrimp_contact_model_draws |> 
  rename(emmean = .value)

shrimp_contact_raw_data <- final_df |> 
  rename(emmean = shrimp_contact)

shrimp_contact_raw_data$stage <- factor(shrimp_contact_raw_data$stage, levels = c("Pre", "During", "Post"))
shrimp_contact_raw_data$stroke <- factor(shrimp_contact_raw_data$stroke, levels = c("control", "4", "2"))

shrimp_contact_model_estimates$stage <- factor(shrimp_contact_model_estimates$stage, levels = c("Pre", "During", "Post"))
shrimp_contact_model_estimates$stroke <- factor(shrimp_contact_model_estimates$stroke, levels = c("control", "4", "2"))

shrimp_contact_means <- shrimp_contact_raw_data |> 
  group_by(stage, stroke) |> 
  summarise(emmean = mean(emmean))

shrimp_contact_fig_together <- ggplot(data = shrimp_contact_raw_data, aes(x = stage, y = emmean, color = stroke, fill = stroke)) +
  geom_point(data = shrimp_contact_means, alpha = 1, position = position_dodgenudge(width = 0.8, x = -0.09), size = 9, shape = "_") +
  geom_half_point(alpha = 0.3, position = position_dodgenudge(width = 0.8, x = -0.03), side = "l", range_scale = 0.5) +
  geom_half_violin(data = shrimp_contact_model_draws, aes(x = stage, y = emmean, color = stroke, fill = stroke), alpha = 0.6, position = position_dodge(width = 0.8), side = "r", size = 0) +
  geom_point(data = shrimp_contact_model_estimates, aes(x = stage, y = emmean, color = stroke), position = position_dodge(width = 0.8), size = 3) +
  geom_errorbar(data = shrimp_contact_model_estimates, aes(x = stage, y = emmean, ymin = lower.HPD, ymax = upper.HPD, color = stroke), position = position_dodge(width = 0.8), width = 0, size = 1.25) +
  scale_fill_manual(values = c("control" = "#76A34A", "4" = "#F29E00", "2" = "#E25A00"), name = "Noise exposure", guide = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = c("control" = "#2A330E", "4" = "#805213", "2" = "#542709"), name = "Noise exposure", guide = guide_legend(override.aes = list(size = 3))) +
  scale_x_discrete(labels = c("Pre", "During", "Post")) +
  labs(title = "The effects of boat noise on goby-shrimp contact",
       y = "Proportion of time the in contact",
       x = "Experimental phase relative to noise exposure") +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 10, margin = margin(t = 5)),  
    axis.text.y = element_text(size = 10, margin = margin(r = 5)),  
    axis.title.x = element_text(size = 13, margin = margin(t = 10)),
    axis.title.y = element_text(size = 13, margin = margin(r = 10)),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    legend.position = "right",
    legend.justification = "top",
    plot.title = element_text(hjust = 0.5, size = 16),  
    plot.margin = margin(10, 10, 10, 10)
  )

shrimp_contact_fig_together
ggsave("./2-figures/shrimp_contact_fig_together.png", plot = shrimp_contact_fig_together, units = "px", width = 2200, height = 1400, dpi = 300)
```

@@

```{r}
shrimp_contact_plot3
ggsave("./2-figures/shrimp_contact_plot3.jpeg", plot = shrimp_contact_plot3, units = "px", width = 2200, height = 1400, dpi = 300)
ggsave("./2-figures/shrimp_contact_plot3.pdf", plot = shrimp_contact_plot3, units = "px", width = 2200, height = 1400)
```

@@

write_csv(model_estimates, "./1-data/model_estimates.csv")
```

@@

write_csv(goby_out_contrasts_table, "./1-data/goby_out_contrasts_table.csv")

@@

write_csv(shrimp_out_contrasts_table, "./1-data/shrimp_out_contrasts_table.csv")

@@

write_csv(shrimp_contact_contrasts_table, "./1-data/shrimp_contact_contrasts_table.csv")
```

@@

>>>>

**SUMMARY**  
Persistent noise pollution produced by boat traffic is reshaping marine soundscapes globally. Despite growing ecological concern, most studies to date have focused on individual-level effects under laboratory conditions, leaving major gaps in our understanding of how boat noise shapes species interactions in the wild. Using field-based behavioural assays, we investigate how boat noise from different engine types (4-stroke and 2-stroke) affects the mutualistic partnership between Steinitz’s goby (Amblyeleotris steinitzi) and snapping shrimp (Alpheus spp.). Across 123 partnerships, we recorded behavioural responses before, during, and after noise exposure. Gobies increased burrow use during 4-stroke boat noise exposure, while shrimp responded stronger to 2-stroke noise—reflecting taxon-specific sensitivities to different noise spectra. Despite these shifts, tactile partner communication remained robust across treatments. These findings highlight divergent vulnerabilities between species tied to different engine acoustics, and emphasise the need for targeted research to inform strategies for mitigating marine noise pollution.

@@

**CONTRIBUTOR ROLES**  
Here, we report a list of the author's roles using the Contributor Role Taxonomy (CRediT). Jack L. Manera: Data curation (lead), Formal analysis (lead), Investigation (contributor), Methodology (contributor), Visualization (lead), Writing – Original Draft Preparation (co-lead), and Writing – Review & Editing (co-lead). Jake M. Martin: Data curation, Formal analysis, Supervision, Methodology, Writing – Original Draft Preparation (co-lead), and Writing – Review & Editing (co-lead). Maria M. Palacios: Conceptualisation, Investigation, Methodology, Writing – Review & Editing. Rachel T. Mason: Data curation, Writing – Review & Editing. Mark I. McCormick: Conceptualisation, Investigation, Methodology, Writing – Review & Editing. Bob BM. Wong: Conceptualisation, Investigation, Methodology, Funding Acquisition, Writing – Review & Editing.

@@

**Licenses/restrictions placed on the data:**
CC BY-NC-SA 4.0 Attribution-NonCommercial-ShareAlike 4.0 International [https://creativecommons.org/licenses/by-nc-sa/4.0/]

**Link to the associated publication:**     
(to be updated upon publication)

**Recommended citation for this data:**      
Manera JL, Martin JM, Palacios MM, Mason RT, McCormick MI, Wong BBM. 2025. Data from: Boat noise alters behaviour in a fish-shrimp mutualism. DOI: 10.17605/OSF.IO/3X5YF. (to be updated upon publication)

@@

@@

```{r}
#transforming the data frame into wide format and turning time spent in each zone into a proportion of the total time (i.e. a value between 0 and 1)
wide_data <- workingdf5 |>
  pivot_wider(
    id_cols = c(burrow_species, burrow_species_stage, burrow, stage, species, boat_id, stroke, site, number_conspecifics, shrimp_species, cum_sum_boat_trials_per_site, date),
    names_from = behaviour,
    values_from = mean_duration
  )

@@

final_df <- inner_join(goby_df, shrimp_df, by = "burrow_stage") |> 
  mutate(shrimp_contact = ifelse(is.na(shrimp_contact), ifelse(goby_out == 1, 0, 1 - goby_out), shrimp_contact)) |> 
  filter(site != "SITE O") |> 
  mutate(parsed_date = mdy(paste(date, "2023")),
         experiment_day = dense_rank(parsed_date))
```

#double check this
```{r}
final_df <- inner_join(goby_df, shrimp_df, by = "burrow_stage") %>% 
  mutate(
    shrimp_contact = ifelse(is.na(shrimp_contact),
                            ifelse(goby_out == 1, 0, 1 - goby_out),
                            shrimp_contact)) %>% 
  filter(site != "SITE O") %>% 
  mutate(
    parsed_date    = mdy(paste(date, "2023")),
    experiment_day = as.integer(parsed_date - min(parsed_date, na.rm = TRUE)) + 1
  )

head(final_df)
```

```{r}
write_csv(final_df, "./1-data/final_df.csv")
```

@@

```{r}
get_prior(zoib_formula, data = final_df)
```

@@

prior_model <- brm(
  prior_formula,
  data = final_df,
  family = zero_one_inflated_beta(),
  prior = priors,
  control = list(adapt_delta = 0.99,
                 max_treedepth = 12),
  chains = 4, iter = 1000, warmup = 100,
  cores = 4, threads = threading(2),
  backend = "cmdstanr",
  seed = 1,
  sample_prior = "only",
  file = "./3-models/prior_check"
)

@@

```{r}
goby_out_model <- brm(
  goby_out_formula_1,
  data = final_df,
  family = zero_one_inflated_beta(),
  prior = priors,
  control = list(adapt_delta = 0.99,
                 max_treedepth = 12),
  chains = 4, iter = 4000, warmup = 1000,
  cores = 4, threads = threading(2),
  backend = "cmdstanr",
  seed = 1,
  file = "./3-models/goby_out_model"
)
```

@@

```{r}
goby_out_plot3
ggsave("./2-figures/goby_out_plot.png", plot = goby_out_plot3, units = "px", width = 2200, height = 1400, dpi = 300)
```

@@

```{r}
shrimp_out_model <- brm(
  shrimp_out_formula,
  data = final_df,
  family = zero_one_inflated_beta(),
  prior = priors,
  control = list(adapt_delta = 0.99,
                 max_treedepth = 12),
  chains = 4, iter = 4000, warmup = 1000,
  cores = 4, threads = threading(2),
  backend = "cmdstanr",
  seed = 1,
  file = "./3-models/shrimp_out_model"
)
```

@@

shrimp_out_plot3
ggsave("./2-figures/shrimp_out_plot.png", plot = shrimp_out_plot3, units = "px", width = 2200, height = 1400, dpi = 300)
```

@@

```{r}
shrimp_contact_model <- brm(
  bf(shrimp_contact ~ stroke*stage + shrimp_number + goby_number + shrimp_species + (1|site/burrow),
     phi ~ stroke*stage + shrimp_number + goby_number + shrimp_species + (1|site/burrow),
     zoi ~ stroke*stage + shrimp_number + goby_number + shrimp_species + (1|site/burrow),
     coi ~ stroke*stage + shrimp_number + goby_number + shrimp_species + (1|site/burrow)
     ),
  data = final_df,
  prior = priors,
  family = zero_one_inflated_beta(),
  control = list(adapt_delta = 0.99, max_treedepth = 12),
  chains = 4, iter = 6000, warmup = 1000, seed = 1, backend = "cmdstanr", cores = 4, threads = threading(2),
  file = "./3-models/shrimp_contact_model"
)
```

@@

shrimp_contact_model_estimates_two <- shrimp_contact_model_estimates |> 
  dplyr::filter(stroke == "2")|> 
  rename(.value = emmean)
```

@@

```{r}
shrimp_contact_plot3
ggsave("./2-figures/shrimp_contact_plot.png", plot = shrimp_contact_plot3, units = "px", width = 2200, height = 1400, dpi = 300)
```

@@

write_csv(model_estimates, "./2-figures/supplementary_tables/Table_S2.csv")
```

@@

write_csv(goby_out_contrasts_table, "./2-figures/supplementary_tables/Table_S3.csv")

@@

write_csv(shrimp_out_contrasts_table, "./2-figures/supplementary_tables/Table_S4.csv")

@@

write_csv(shrimp_contact_contrasts_table, "./2-figures/supplementary_tables/Table_S5.csv")
```

@@

# Supplementary analysis

Although sites in the experimental design were spatially distinct, boats were tested multiple times at the same sites across the experiment. This raises the possibility that some burrows may have potentially experienced prior boat-noise exposure before their focal trial. Such cumulative exposure could influence the behavioural responses we observed. To assess this, we conducted a supplementary analysis.

We quantified potential cumulative prior exposure at each site using the variable cum_sum_boat_trials_per_site, defined as the cumulative number of boat-noise treatments that had occurred at a site before the current trial. This variable was added as a predictor to the original models. We then compared models with and without the predictor using LOOIC, examined the estimated trend of each behavioural outcome in relation to cumulative exposure, and visually and quantitatively compared effect sizes. Finally, we assessed whether inclusion of the cumulative exposure variable altered the inference of the main effects.

```{r}
ggplot(data = final_df, aes(x = cum_sum_boat_trials_per_site, y = goby_out)) +
  geom_smooth(method = "lm") +
  geom_jitter()

ggplot(data = final_df, aes(x = cum_sum_boat_trials_per_site, y = shrimp_out)) +
  geom_smooth(method = "lm") +
  geom_jitter()

ggplot(data = final_df, aes(x = cum_sum_boat_trials_per_site, y = shrimp_contact)) +
  geom_smooth(method = "lm") +
  geom_jitter()
```

## Goby out model

### Fit model

Here we run the models with 'cum_sum_boat_trials_per_site' which is the cumulative sum of the number of boat trials at each site. This is to see if potential prior exposure to boat noise has a an effect on behavior

```{r}
cumulative_sum_goby_out_formula <- bf(
  goby_out ~ stroke*stage + goby_number + shrimp_number + shrimp_species + cum_sum_boat_trials_per_site + (1 | site/burrow),
  phi ~ stroke*stage + goby_number + shrimp_number + shrimp_species + cum_sum_boat_trials_per_site + (1 | site/burrow),
  zoi ~ stroke*stage + goby_number + shrimp_number + shrimp_species + (1 | site/burrow),
  coi ~ stroke*stage + goby_number + shrimp_number + shrimp_species + (1 | site/burrow),
  family = zero_one_inflated_beta()
  )

cumulative_sum_goby_out_model <- brm(
  cumulative_sum_goby_out_formula,
  data = final_df,
  family = zero_one_inflated_beta(),
  prior = priors,
  control = list(adapt_delta = 0.99,
                 max_treedepth = 12),
  chains = 4, iter = 4000, warmup = 1000,
  cores = 4, threads = threading(2),
  backend = "cmdstanr",
  seed = 1,
  file = "./3-models/cumulative_sum_goby_out_model"
)
```

### Model Diagnostics

```{r}
summary(cumulative_sum_goby_out_model)

plot(cumulative_sum_goby_out_model, ask = F)

pp_check(cumulative_sum_goby_out_model)
```

### Model predictive performance

```{r}
goby_out_model <- add_criterion(goby_out_model, criterion = "loo")
cumulative_sum_goby_out_model <- add_criterion(cumulative_sum_goby_out_model, criterion = "loo")

loo_compare(goby_out_model, cumulative_sum_goby_out_model)
```

the addition of 'cum_sum_boat_trials_per_site' does not improve model predictive performance

### Effects of previous boat trials

```{r}
emtrends(cumulative_sum_goby_out_model, ~ 1, var = "cum_sum_boat_trials_per_site", epred = TRUE, re_formula = NA)
```

'cum_sum_boat_trials_per_site' does not have a meaningful effect on the proportion of time gobies spent out of the burrow

### Comparison of effect estimates

```{r}
cumulative_sum_goby_out_model_contrast_effect <- cumulative_sum_goby_out_model |>
  emmeans(~ stroke*stage, 
          epred = TRUE,
          re_formla = NULL,
          rg.limit =14000) |> 
  contrast(method = "revpairwise")

# Extract the relevant contrast information
cumulative_sum_goby_out_results.table <- as.data.frame(cumulative_sum_goby_out_model_contrast_effect) |> 
  dplyr::filter(contrast %in% c("control Pre - 2 Pre", "control Pre - 4 Pre","4 Pre - 2 Pre", "control Pre - control During", "control Pre - control Post", "4 Pre - 4 During", "4 Pre - 4 Post", "2 Pre - 2 During", "2 Pre - 2 Post"))

cumulative_sum_goby_out_results.table
goby_out_results.table

cumulative_sum_goby_out_results.table$dataset <- "Cumulative sum model"
goby_out_results.table$dataset <- "Non-cumulative model"

goby_contrast_plot_data <- bind_rows(
  cumulative_sum_goby_out_results.table,
  goby_out_results.table
)

# Plot contrasts with HPD intervals
ggplot(goby_contrast_plot_data,
       aes(x = estimate, y = contrast, color = dataset)) +
  geom_point(position = position_dodge(width = 0.6), size = 2) +
  geom_errorbarh(aes(xmin = lower.HPD, xmax = upper.HPD),
                 position = position_dodge(width = 0.6), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  theme_classic() +
  labs(
    x = "Posterior median (with 95% HPD interval)",
    y = "Contrasts",
    title = "Comparison of contrasts between cumulative-sum and original models"
  )
```

## Shrimp out model

### Fit model

Here we run the models with 'cum_sum_boat_trials_per_site' which is the cumulative sum of the number of boat trials at each site. This is to see if potential prior exposure to boat noise has a an effect on behavior

```{r}
cumulative_sum_shrimp_out_formula <- bf(
  shrimp_out ~ stroke*stage + goby_number + shrimp_number + shrimp_species + cum_sum_boat_trials_per_site + (1 | site/burrow),
  phi ~ stroke*stage + goby_number + shrimp_number + shrimp_species + cum_sum_boat_trials_per_site + (1 | site/burrow),
  zoi ~ stroke*stage + goby_number + shrimp_number + shrimp_species + (1 | site/burrow),
  coi ~ stroke*stage + goby_number + shrimp_number + shrimp_species + (1 | site/burrow),
  family = zero_one_inflated_beta()
  )

cumulative_sum_shrimp_out_model <- brm(
  cumulative_sum_shrimp_out_formula,
  data = final_df,
  family = zero_one_inflated_beta(),
  prior = priors,
  control = list(adapt_delta = 0.99,
                 max_treedepth = 12),
  chains = 4, iter = 4000, warmup = 1000,
  cores = 4, threads = threading(2),
  backend = "cmdstanr",
  seed = 1,
  file = "./3-models/cumulative_sum_shrimp_out_model"
)
```

### Model Diagnostics

```{r}
summary(cumulative_sum_shrimp_out_model)

plot(cumulative_sum_shrimp_out_model, ask = F)

pp_check(cumulative_sum_shrimp_out_model)
```

### Model predictive performance

```{r}
shrimp_out_model <- add_criterion(shrimp_out_model, criterion = "loo")
cumulative_sum_shrimp_out_model <- add_criterion(cumulative_sum_shrimp_out_model, criterion = "loo")

loo_compare(shrimp_out_model, cumulative_sum_shrimp_out_model)
```

the addition of 'cum_sum_boat_trials_per_site' does not improve model predictive performance

### Effects of previous boat trials

```{r}
emtrends(cumulative_sum_shrimp_out_model, ~ 1, var = "cum_sum_boat_trials_per_site", epred = TRUE, re_formula = NA)
```

'cum_sum_boat_trials_per_site' does not have a meaningful effect on the proportion of time shrimp spent out of the burrow

### Comparison of effect estimates

```{r}
cumulative_sum_shrimp_out_model_contrast_effect <- cumulative_sum_shrimp_out_model |>
  emmeans(~ stroke*stage, 
          epred = TRUE,
          re_formla = NULL,
          rg.limit =14000) |> 
  contrast(method = "revpairwise")

# Extract the relevant contrast information
cumulative_sum_shrimp_out_results.table <- as.data.frame(cumulative_sum_shrimp_out_model_contrast_effect) |> 
  dplyr::filter(contrast %in% c("control Pre - 2 Pre", "control Pre - 4 Pre","4 Pre - 2 Pre", "control Pre - control During", "control Pre - control Post", "4 Pre - 4 During", "4 Pre - 4 Post", "2 Pre - 2 During", "2 Pre - 2 Post"))

cumulative_sum_shrimp_out_results.table
shrimp_out_contrasts

cumulative_sum_shrimp_out_results.table$dataset <- "Cumulative sum model"
shrimp_out_contrasts$dataset <- "Non-cumulative model"

shrimp_contrast_plot_data <- bind_rows(
  cumulative_sum_shrimp_out_results.table,
  shrimp_out_contrasts
)

# Plot contrasts with HPD intervals
ggplot(shrimp_contrast_plot_data,
       aes(x = estimate, y = contrast, color = dataset)) +
  geom_point(position = position_dodge(width = 0.6), size = 2) +
  geom_errorbarh(aes(xmin = lower.HPD, xmax = upper.HPD),
                 position = position_dodge(width = 0.6), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  theme_classic() +
  labs(
    x = "Posterior median (with 95% HPD interval)",
    y = "Contrasts",
    title = "Comparison of contrasts between cumulative-sum and original models"
  )
```

## Shrimp contact model

### Fit model

Here we run the models with 'cum_sum_boat_trials_per_site' which is the cumulative sum of the number of boat trials at each site. This is to see if potential prior exposure to boat noise has a an effect on behavior

```{r}
cumulative_sum_shrimp_contact_formula <- bf(
  shrimp_contact ~ stroke*stage + goby_number + shrimp_number + shrimp_species + cum_sum_boat_trials_per_site + (1 | site/burrow),
  phi ~ stroke*stage + goby_number + shrimp_number + shrimp_species + cum_sum_boat_trials_per_site + (1 | site/burrow),
  zoi ~ stroke*stage + goby_number + shrimp_number + shrimp_species + (1 | site/burrow),
  coi ~ stroke*stage + goby_number + shrimp_number + shrimp_species + (1 | site/burrow),
  family = zero_one_inflated_beta()
  )

cumulative_sum_shrimp_contact_model <- brm(
  cumulative_sum_shrimp_contact_formula,
  data = final_df,
  family = zero_one_inflated_beta(),
  prior = priors,
  control = list(adapt_delta = 0.99,
                 max_treedepth = 12),
  chains = 4, iter = 4000, warmup = 1000,
  cores = 4, threads = threading(2),
  backend = "cmdstanr",
  seed = 1,
  file = "./3-models/cumulative_sum_shrimp_contact_model"
)
```

### Model Diagnostics

```{r}
summary(cumulative_sum_shrimp_contact_model)

plot(cumulative_sum_shrimp_contact_model, ask = F)

pp_check(cumulative_sum_shrimp_contact_model)
```

### Model predictive performance

```{r}
shrimp_contact_model <- add_criterion(shrimp_contact_model, criterion = "loo")
cumulative_sum_shrimp_contact_model <- add_criterion(cumulative_sum_shrimp_contact_model, criterion = "loo")

loo_compare(shrimp_contact_model, cumulative_sum_shrimp_contact_model)
```

the addition of 'cum_sum_boat_trials_per_site' does not improve model predictive performance

### Effects of previous boat trials

```{r}
emtrends(cumulative_sum_shrimp_contact_model, ~ 1, var = "cum_sum_boat_trials_per_site", epred = TRUE, re_formula = NA)
```

'cum_sum_boat_trials_per_site' does not have a meaningful effect on the proportion of time shrimp spent in contact with a goby

### Comparison of effect estimates

```{r}
cumulative_sum_shrimp_contact_model_contrast_effect <- cumulative_sum_shrimp_contact_model |>
  emmeans(~ stroke*stage, 
          epred = TRUE,
          re_formla = NULL,
          rg.limit =14000) |> 
  contrast(method = "revpairwise")

# Extract the relevant contrast information
cumulative_sum_shrimp_contact_results.table <- as.data.frame(cumulative_sum_shrimp_contact_model_contrast_effect) |> 
  dplyr::filter(contrast %in% c("control Pre - 2 Pre", "control Pre - 4 Pre","4 Pre - 2 Pre", "control Pre - control During", "control Pre - control Post", "4 Pre - 4 During", "4 Pre - 4 Post", "2 Pre - 2 During", "2 Pre - 2 Post"))

cumulative_sum_shrimp_contact_results.table
shrimp_contact_contrasts

cumulative_sum_shrimp_contact_results.table$dataset <- "Cumulative sum model"
shrimp_contact_contrasts$dataset <- "Non-cumulative model"

shrimp_contact_contrast_plot_data <- bind_rows(
  cumulative_sum_shrimp_contact_results.table,
  shrimp_contact_contrasts
)

# Plot contrasts with HPD intervals
ggplot(shrimp_contact_contrast_plot_data,
       aes(x = estimate, y = contrast, color = dataset)) +
  geom_point(position = position_dodge(width = 0.6), size = 2) +
  geom_errorbarh(aes(xmin = lower.HPD, xmax = upper.HPD),
                 position = position_dodge(width = 0.6), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  theme_classic() +
  labs(
    x = "Posterior median (with 95% HPD interval)",
    y = "Contrasts",
    title = "Comparison of contrasts between cumulative-sum and original models"
  )
```
